Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 paper.tex
Rewrite GPU intro para
Closes #249

\subsection{Future applications: GPU acceleration}
GPU acceleration has revolutionized scientific computing by leveraging the
massively parallel architecture of GPUs to achieve significant speedups in
computationally intensive tasks such as simulations, data analysis, and machine
learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
mittal2014survey,owens2008gpu}
% (Nickolls & Dally, 2010;
% Mittal & Vetter, 2015; 
% Owens et al., 2008). 
This has profound implications for
fields ranging from molecular dynamics~\citep{stone2007accelerating}
% (Stone et al., 2010) 
to deep learning~\citep{lecun2015deep}
% (LeCun et al., 2015). 
While early GPU programming required expertise in
low-level languages like CUDA or OpenCL, the advent of high-level Python
libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
%(Okuta et al., 2017), 
and
PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
% (Klöckner, 2012) 
has democratized access to GPU acceleration.
These tools offer user-friendly interfaces and just-in-time compilation,
allowing scientists to harness the power of GPUs within the familiar Python
ecosystem, thereby significantly accelerating scientific discovery and
broadening the adoption of accelerated computing across various research
domains.

% [JK: this is a super rough first narrative draft.]
% Machine learning, and in particular deep learning methods, have led to
% major breakthroughs in many different fields in recent years [CITES].
% GPUs have been a key enabling factor in this [details and cites].
% The PyData ecosystem [cites] has also enabled this. Techologies like
% [Cupy, lots of others] substantially lower the bar for writing efficient
% GPU code.
% Deep learning methods depend crucially on efficient access to vast
% datasets for training, and Zarr is emerging as a popular choice
% [cites from geosciences?]. The RAPIDs GPU Accelerated data science
% library [which does] now supports Zarr through [kvikio, more?].

@@

There is therefore an opportunity to develop a new 
generation of Zarr-native applications, taking advantage of the efficient data
representation and natural model of chunk-based parallelism.
Although there are several independent Zarr implementations
across programming languages (see Methods),
the Python data science ecosystem is a particularly powerful
platform,
with a rich suite of tools~\cite[e.g.][]{
mckinney2010data,
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy}
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Xarray~\cite{hoyer2017xarray} provides a unified interface for working
with multi-dimensional arrays
in Python, and libraries like Dask~\cite{rocklin2015dask}
and Cubed~\cite{cubed2024} allow these operations to be
scaled out transparently across processors and clusters.
This scaling is achieved by distributing calculations over
grid-based array representations like Zarr, where chunks
provide the basic unit for parallel computation.
The VCF Zarr specification introduced here was created
to facilitate work on a scalable genetics toolkit for
Python~\cite{sgkit2024} built on Xarray.
While the high-level facilities for distributed
computation provided by Xarray
are very powerful, they are not needed or indeed
appropriate in all contexts. Our benchmarks here
illustrate that working at the lowest level,
by applying optimised kernels on a chunk-by-chunk
basis is both straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers
can build utilities using the VCF Zarr specification
using the appropriate level of abstraction and tool chain
on a case-by-case basis.

>>>>

\subsection{Future applications: GPU acceleration}
GPU acceleration has revolutionized scientific computing by leveraging 
their massively parallel architecture to achieve significant speedups in
computationally intensive tasks such as simulations, data analysis, and machine
learning, often by one or two orders of magnitude~\citep{nickolls2010gpu,
mittal2014survey,owens2008gpu}
This has had a transformative impact across many 
% TODO More
fields~\citep[e.g.][]{stone2007accelerating,pandey2022transformational}.
While early GPU programming required expertise in
low-level languages like CUDA or OpenCL, the advent of high-level Python
libraries such as Numba~\citep{lam2015numba}, CuPy~\citep{okuta2017cupy} 
and PyCUDA/PyOpenCL~\citep{klockner2012pycuda}
has made GPU acceleration far more accessible.
Deep learning~\citep{lecun2015deep} has led to major breakthroughs in 
several domains in recent 
% TODO more?
years~\citep[e.g.][]{jumper2021highly,merchant2023scaling}
and depends crucially on processing large volumes of data
on GPUs or other custom hardware accelerators.
The Python data science ecosystem is particularly powerful in this area,
with GPU-native libraries like
PyTorch~\cite{paszke2019pytorch}
and TensorFlow~\citep{tensorflow2015-whitepaper}
providing the basic infrastructure for the field.

@@

There is therefore an opportunity to develop a new 
generation of Zarr-native applications, taking advantage of the efficient data
representation and natural model of chunk-based parallelism.
Although there are several independent Zarr implementations
across programming languages (see Methods),
the Python data science ecosystem is a particularly powerful
platform,
with a rich suite of tools~\cite[e.g.][]{
mckinney2010data,
scikit-learn, 
lam2015numba,
kluyver2016jupyter,
harris2020array,
virtanen2020scipy}
and is increasingly
popular in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Xarray~\cite{hoyer2017xarray} provides a unified interface for working
with multi-dimensional arrays
in Python, and libraries like Dask~\cite{rocklin2015dask}
and Cubed~\cite{cubed2024} allow these operations to be
scaled out transparently across processors and clusters.
This scaling is achieved by distributing calculations over
grid-based array representations like Zarr, where chunks
provide the basic unit for parallel computation.
The VCF Zarr specification introduced here was created
to facilitate work on a scalable genetics toolkit for
Python~\cite{sgkit2024} built on Xarray.
While the high-level facilities for distributed
computation provided by Xarray
are very powerful, they are not needed or indeed
appropriate in all contexts. Our benchmarks here
illustrate that working at the lowest level,
by applying optimised kernels on a chunk-by-chunk
basis is both straightforward to implement and highly performant.
Thus, a range of possibilities exist in which developers
can build utilities using the VCF Zarr specification
using the appropriate level of abstraction and tool chain
on a case-by-case basis.
